Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-12-11 demystify.tex
updates

\mainmatter
\chapter{Should You Apply to a CS PhD Program in the US?}\label{sec:should}

@@

However, MS can help admission if it gives research experience or is from a more well-known school than your undergrad institution (\autoref{sec:your-school}).
Moreover, if you have an MS then some course work \emph{might be} transferred for course credits, which \emph{might} save some time. But in general don't count finishing earlier just because you have an MS.

@@

\chapter{Your Application}\label{sec:application}

\chapterinfo{The committee will look at various factors, but the most important ones are letters of recommendation (LORs),  statements of purpose (SOP), and research experience and potential, e.g., publications.}

@@

The goal of adcom is to evaluate your research experience, potential, and interest to see if you \textbf{fit into our PhD program}! The emphasis here is \emph{fitting}, which varies from school to school, faculty to faculty, and even from year to year.  However, in general, to evaluate your profile, we consider the following key indicators.

\section{Letters of Recommendation (LOR)}\label{sec:lor}
\sectioninfo{\emph{LORs are very important}, but only if they are personalized and talk about your research ability.}

@@

Most PhD programs will require at least \textbf{two LORs}. 
The goal of the LoR is to provide the adcom with an assessment of your research ability and potential. 
The letter should 
focus on your research background, achievements, and potentials, all of which paint a compelling picture of the candidate to the adcom.

\paragraph{LoR writers} LoR writers should be someone who (i) can talk in depth about your research experience and potential and (ii) have the credibility to evaluate your research ability.

A LoR writer should be someone who \emph{knows you well}, so that they can provide a strong-supporting and personalized letter for you.  LoR writers should also have the credibility to convince adcom members on your research ability and potential. Such a person is typically someone who has a PhD (so they know what a PhD is about) and is active in research (so they can evaluate your research ability). LoR writers are often your research advisers, professors, or supervisors who have worked with you on research projects.

\paragraph{LoR from Famous People}  Having a strong letter from an internationally recognized researcher will \emph{greatly strengthen} your application. A strong LOR from such a researcher can outweigh the lack of publications (or low GPA or unknown school). The reason is simple: adcom members trust people that they know and respect. If a well-known researcher enthusiastically recommends you, then the adcom members will likely trust that recommendation.

@@

However, let's face it.  Most students don't have an opportunity to interact or work with well-known researchers.  
So what to do? It is perfectly fine to have a letter from people who know you well enough to talk about \emph{your research experience and capabilities}. These include profs., researchers, postdocs who have mentored and worked with you. In fact, a personal, customized, and strong letter from even a senior PhD grad student who mentors and knows you well is much better than a generic letter from a well-known superstar who does not know you well.

@@

Sometimes students would go to great lengths just to get letters from well-known professors in their school (e.g., department head or dean). But as mentioned, if these professors do not know you, their letters would likely be generic and carry little value (sometimes \red{red flags}). Moreover, a top professor at your university might not be well-known to US faculty (see more details in \autoref{sec:admin-letters} and \autoref{sec:your-school}). So save the trouble and get letters from \emph{any} professors/supervisors who know you well and can write a good letter about \emph{your} research ability. It's better to have a good personalized letter about your own research ability from someone who is less well-known than a generic/weak letter from a well-known person.

@@

\subsection{Self-written Letters are Bad}\label{sec:self-letters}
Many students write letters themselves, typically due to the requests of their profs. or letter writers. Such letters have \emph{little value} and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the ref. writer talking about you, in this case it is you who write about yourself (and they just sign the letter).

@@

\subsection{What can you do?}\label{sec:lor-what-to-do}

\paragraph{Waiving Your Right} When you ask someone to write a letter for you, you will be asked if you want to waive your right to see the letter.  You should \emph{always} waive your right.

@@

\paragraph{Help your LOR Writer} As mentioned above in \autoref{sec:self-letters} and \autoref{sec:generic-letters}, do not write your own letter and generic letters do not give much value. Thus, to help your writer to write a strong, customized LoR, you should provide them details or unique things about yourself. For example, let them know about your GPA, research and work experience, papers (if any), or anything you want them to mention.  If the GPA in your program is highly competitive (\autoref{sec:gpa}) and they know that, remind them to talk about it in the LOR. 
You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and complement that with their own perspective.

@@

\section{Research Experience}\label{sec:research-experience}

@@

This section discusses publications and other research experiences that can strengthen your application.
\autoref{sec:research-opportunities} provides more information on how to find research opportunities (e.g., during your undergrad study).

\subsection{Publications} The most concrete evidence of research ability is having \emph{papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that you have been successfully involved in research.

@@

\begin{commentbox}[Thanh:]
    Due to academic culture, professors in Vietnam usually aim for (international) journals instead of conferences. Could you give some tips on how to know whether a journal is good (CSrankings, unfortunately, only consider conferences)?
    \tcblower
    \textbf{Vu}: One way is looking at what well-known researchers publish. For example, if you are interested in a field X, you can use CSRankings to look at active faculty in X, and then look at their websites to see what journals they publish at.
\end{commentbox}
\paragraph{What if I don't have any pubs} It is understandable that many students do not have the opportunity to publish papers. Thus, other writings, even those under submissions or rejected, would still be good and better than nothing.  Be sure to upload your papers with your application and talk about them in your statement (see \autoref{sec:sop}).

@@

\subsection{Work Experience} Work experience at \emph{well-known research laboratories}, such as Microsoft Research, can strengthen your
application.  Note that the emphasis here is \emph{research} places, not software development or non-research work. For example, working at a FAANG (Facebook, Amazon, Apple, Netflix, Google) company as a software engineer is not research experience, though it can be helpful for MS applications discussed in \autoref{chap:ms}. Similarly, a LoR from your supervisor for non-research experience might not count much. So do not spend much time talking about non-research job in your SOP.

Note that adcom reviewers might not be familiar with all research labs, especially those outside the US For example, while VinAI is  well-known in Vietnam and potentially in Asia, it might not be very well-known in the US. So you or your LoR writers should explicitly say something about them in your statement.  In general, if you did some good research work, then you should mention that in your SOP and ask your supervisor to write about it in their LOR (\autoref{sec:lor}).

@@

\subsection{Competitions} Winning \emph{internationally recognized competitions} can demonstrate your research potential.
For example, participating in Math Olympiads if you want to do theory or winning ACM programming contests if you want to ``build'' systems, e.g., software analysis.

\section{Statement of Purpose (SOP)}\label{sec:sop}
\sectioninfo{\emph{SOP is important}. Write it in such a way that makes you \emph{stand out}.}

@@

While you might not have control over LORs (\autoref{sec:lor}) or where your go to school (\autoref{sec:your-school}), you do over your
statement of purpose (SOP) or personal statement\footnote{Few schools separate these documents and ask you to write both: SOP, which focuses on research experiences, and personal statement, which is everything more personal, e.g., why PhD, challenges, diversity, etc}! A well-written SOP also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \autoref{sec:funding}).  Many SOP samples for CS are \href{https://cs-sop.org/}{available here}.

@@

\subsection{Kiss of Death in SOP} \label{sec:kiss-of-death-sop}

@@

\subsection{Diversity Statement}\label{sec:diversity-statement}

@@

\section{Your School and Grades}\label{sec:your-school}
\sectioninfo{\emph{High grades probably won't help much} (unless you're from a very top and well-known school), but bad ones likely will raise concerns.}

\subsection{School} Graduating from top universities \emph{that adcom members recognize helps}. For example, if you are an international student and your school is well-known, then it is \emph{``top foreign''}, which is a plus.
However, if we do not know much about schools in your country, then we are uncertain about the quality of your school and likely treat your school as \emph{``unknown foreign''}, which can be a minus point.

@@

\begin{commentbox}[Vu:]
  Sometimes PhD adcom in the US will share a document such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Foreign-Top-Schools}{this one}, which lists the top schools in several countries. We also ask other faculty and students if we think they know about the place.  For example, when I was a postdoc at UMD, members of their CS PhD adcom asked me to evaluate applicants from Vietnam.  During my time at UNL and now here at Mason, I have looked at Vietnamese applications (whether they are assigned to me or not) and provided input to their reviewers, e.g., X is the top tech school in Vietnam and so it should be \emph{top} instead of \emph{unknown} foreign, which makes a huge difference.
\end{commentbox}
\begin{commentbox}[Deepak:]
  If an applicant is anxious about their school not being known outside their country, they can provide information about their school and department, with independent sources where such information can be verified.
\end{commentbox}
\subsection{Grades}\label{sec:gpa}

Compared to other factors such as LoRs (\autoref{sec:lor}) and research experiences (\autoref{sec:research-experience}), grades generally do not matter much for CS PhD admission. PhD in CS is a research degree and doing well in undergrad courses does not necessarily mean you can do research (other factors such as research experiences, pubs, LoRs are more important).  In fact, many CS faculty members themselves have bad grades in undergrad courses (and some were proud of that!).

Nonetheless, if you are from a well-known school (\autoref{sec:your-school}), having good grades do help (not a lot though), e.g., adcom members often note details such as "good GPA from well-known school (\autoref{sec:ievaluate}). However if your school is not well-known, having top grades or rankings usually will not help because we cannot evaluate them (e.g., we don't know how hard it is to get a 4.0 or A's at your school). This can be an issue for students in many top international universities where the competition is so high that very good students can still have low rankings from these schools (and be overlooked by adcom).

@@

\section{Standard Tests (e.g., GRE and IELTS), Online Courses, and Certificates}\label{sec:standard-tests}
\sectioninfo{\emph{Standard tests are not important}. GRE typically \emph{is not} required. For standard English tests (not required for domestic students), just do enough to pass the minimum requirements.}

\subsection{GREs Are Optional and Do Not Matter} While a few schools still require taking the GRE (e.g., UCF), most good CS PhD programs in the US \textbf{no longer} require it. Thus, if you haven't taken the GRE or have bad scores, then don't waste time (re)taking it. Being optional really means optional, and not taking it will not hurt your application.

@@

\subsection{English Tests (IELTS, TOEFL)} Unless your degrees are from specific countries such as \href{https://www.gmu.edu/international/english-language-requirements}{these}, you will need to take standardized English tests. On one hand you will need to show some level of English proficiency, but on the other hand, you do not need to have very good scores in these tests (many adcom members themselves were once international students and struggled with English).
You should just do well enough to pass the minimum requirement set by the university. %, which is required for TA (\autoref{sec:funding}).
Just as with grades (\autoref{sec:your-school}) and GRE, having high scores in these tests might not help, but having too low scores can be a \red{red flag} and sometimes results in an automatic rejection (\autoref{sec:why-rejected}), e.g., below the minimum requirement.

@@

\subsection{Online Courses and Certificates}

@@

\section{CV/Resume}
\sectioninfo{Highlight and summarize major achievements, e.g., Publications, Programming Awards}

\begin{commentbox}[Vu:]
    Unlike a job application, CV is not as important for PhD admission. We care more about your LOR, SOP, etc.  So do not spend too much time on your CV, just make sure it is easy to skim through and well-organized around research activities and achievements.
\end{commentbox}

\section{Interview and the Waiting Game}\label{sec:interview}
\sectioninfo{Getting an interview is typically a \emph{good sign}; but no interview does not mean rejection}

@@

\subsection{Interviews}

@@

\subsection{Not Getting Interviews}\label{sec:no-interview}
While it is generally good to get an interview, not getting one \emph{does not} mean you're out.  Many programs do not have the tradition of interviewing applicants. For example, at Mason, most admitted students with TA (\autoref{sec:ta}) do not go through interviews.

@@

\subsection{Notification Timeline: Why rejection letters are sent so late?}\label{sec:late-rejection}
\subsectioninfo{Grad programs often wait for the accepted students to make their decisions, typically by April 15, before sending out rejection letters.}

@@

Not much you can do other than to be patient and wait. Do not send emails asking about interviews or status (unless you have interviewed specifically with someone (\autoref{sec:interview}) then you can ask that person for status updates and other questions (\autoref{sec:accept-postpone-decline}).

@@

% You can then share this sheet with your reference writers and remind them to submit LoRs on your behalf (see \autoref{sec:lor}). Also, you should periodically update your writers with your status; a simple note such as \emph{``I am getting interviewed by prof. X at school Y. Do you have any advice?''} can get you a lot of help.

@@

\chapter{Funding}\label{sec:funding}
\chapterinfo{TAs, RAs, and fellowships are main funding sources for PhDs.  TAs are provided by the department to help with classes. RAs are given by profs. to help with their research.  Fellowships, provided by the university, department, or external sources such as government or industry, give move flexibility but can be very competitive.}

@@

%These questions are not useful and should be avoided. First, no one knows! Even adcom members can't tell if you would get admitted to their programs or not. As mentioned throughout this guide, the admission process is complex and involves many factors that are not part of your profile, e.g., whether the faculty have funding or are interested in your background. Second, answers to this questions are often not helpful to you.  Faculty and people knowledgeable about the process will not answer these questions, and those who do are often not knowledgeable and might give you false hope or despair. OK, someone say your GRE is low (which does not even matter for most schools (\autoref{sec:standard-tests}) , are you going to retake it?  Or someone says your GPA is low, are you going to apply to a different school?  No, you should apply to the schools you want to go to and see what happens.  If you don't get in, then you can ask for feedback and improve your profile for the next round.

@@

For each application, our system compiles a single PDF file, which consists of a summary (degrees, GPAs, etc), transcripts, test scores, LoRs, a CV, SOP, and writing samples. I usually read in this order. I start with the \emph{summary}, checking for low GPAs or test scores below the university minimum (\autoref{sec:standard-tests}). I then skim the \emph{transcripts} for low grades in relevant courses, noting issues like ``many low grades in main courses or unknown international school with good GPA''.  These are not as important as LORs or SOP but I read them first because the review system has questions about them (e.g., ``is GPA good? is IELTS sufficient?'')

@@

  \item[LoR] Letter of Recommendation. A letter written by a professor or supervisor that assesses your qualifications and potential for graduate study (\autoref{sec:lor}).

@@

  \item[SOP] Statement of Purpose is a document written by yourself to explains your research interests, background, and reasons for applying to a PhD program (\autoref{sec:sop}).

@@

  \item[TOEFL/IELTS] Test of English as a Foreign Language. A standardized test to measure the English language ability of non-native speakers.
  \item[GRE] Graduate Record Examination. A standardized test that is an admissions requirement for many graduate schools in the US. However, it is not required for most CS PhD programs (\autoref{sec:standard-tests}).

@@

\paragraph{School \autoref{sec:your-school}} Adcom already knows about your school, which is a plus (see \autoref{sec:your-school}). You are also more familiar with the US education system and culture (so hopefully you are aware of the cultural issues listed in \autoref{sec:cultural}).

\paragraph{Standard Tests \autoref{sec:standard-tests}} You do not need to take TOEFL or IELTS because you already did your undergrad (or MS) at a US university.  You might also be more comfortable communicating in English, e.g., contacting professors (\autoref{sec:contact}).

@@

\paragraph{Research Experience \autoref{sec:research-opportunities}} You might have many opportunities to do research as an undergraduate, e.g., through REUs and internships at your undergrad university.  Highlight such experience in your application.

@@

\subsection{Funding}\label{sec:ms-funding}
Unlike PhD programs, which often have funding (\autoref{sec:funding}), MS students are typically not funded. The reason is because MS students do not focus on research and thus are not funded through RA and PhD students have priority for TA positions. 
\begin{itemize}
\item[\textbf{RA}] Profs. are not willing to take MS students as RAs because they are not around long enough to be productive. It can take a while for a student to get up to speed and start being productive, and by that time, they are about to graduate. Moreover, the goal of most MS students is to get a job, not to do research, so they are not as motivated to do research.

\item[\textbf{TA}] MS students are typically not given TA positions because PhDs are given priority. Many depts. do not have enough TAs for PhD students and so they cannot afford to give them to MS students.
\end{itemize}

However, there are always exceptions, e.g., if you have a strong background and can demonstrate that you can be productive in research, then a prof. might be willing to take you as their RA.
Moreover, some schools have TAs for MS students, and you can apply for these positions. For example, Mason CS has quite a few TAs for MS students; my courses in the past two years have MS TAs.

@@

\item \textbf{Living Cost:} MS students are typically not funded, so living cost can be a big factor. You should consider the tuition and living expenses of the area. Note that sometimes living costs, e.g., renting, seem scary at first but students often find way to make it work, e.g., by sharing an apartment with other students.

@@

\chapter{Research Opportunities}\label{sec:research-opportunities}
\chapterinfo{How to get research experience as an undergrad.}

@@

By contributing code, fixing bugs, implementing new features, or providing documentation, students can gain practical research experience and interact with experienced developers and researchers in the field. Not only you gain research experience, but you might be able to get a LoR from the project maintainer, and you should write about this experience in your SOP (\autoref{sec:sop}.)

@@

>>>>

\mainmatter

\part{Introduction To CS PhD Admission in the US}\label{part:intro}

\chapter{Should You Apply to a CS PhD Program in the US?}\label{sec:should}

@@

However, MS can help admission if it gives research experience or is from a more well-known school than your undergrad institution (\autoref{chap:your-school}).
Moreover, if you have an MS then some course work \emph{might be} transferred for course credits, which \emph{might} save some time. But in general don't count finishing earlier just because you have an MS.

@@

\part{Your Application Materials}\label{part:application}

%\partinfo{.}

@@

The goal of adcom is to evaluate your research experience, potential, and interest to see if you \textbf{fit into our PhD program}! The emphasis here is \emph{fitting}, which varies from school to school, faculty to faculty, and even from year to year.  
The committee will look at various factors, but the most important ones are letters of recommendation (LORs),  statements of purpose (SOP), and research experience and potential, e.g., publications.

\chapter{Letters of Recommendation (LOR)}\label{chap:lor}
\chapterinfo{\emph{LORs are very important}, but only if they are personalized and talk about your research ability.}

@@

Most PhD programs will require at least \textbf{two LORs}. 
The goal of the LoR is to provide the adcom with an assessment of your research ability and potential. 
The letter should 
focus on your research background, achievements, and potentials, all of which paint a compelling picture of you as a researcher.

\section{LoR writers}\label{sec:lor-writers}

LoR writers should be someone who (i) can talk in depth about your research experience and potential and (ii) have the credibility to evaluate your research ability.

A LoR writer should be someone who \emph{knows you well}, so that they can provide a strong-supporting and personalized letter for you.  LoR writers should also \emph{have the credibility} to convince adcom members on your research ability and potential. Such a person is typically someone who has a PhD (so they know what a PhD is about) and is active in research (so they can evaluate your research ability). LoR writers are often your research advisers, professors, or supervisors who have worked with you on research projects.

\subsection{LoR from Famous People}  Having a strong letter from an internationally recognized researcher will \emph{greatly strengthen} your application. A strong LOR from such a researcher can outweigh the lack of publications (or low GPA or unknown school). The reason is simple: adcom members trust people that they know and respect. If a well-known researcher enthusiastically recommends you, then the adcom members will likely trust that recommendation.

@@

However, most students don't have an opportunity to interact or work with well-known researchers.  
So what to do? It is perfectly fine to have a letter from people who know you well enough to talk about \emph{your research experience and capabilities}. These include profs., researchers, postdocs who have mentored and worked with you. In fact, a personal, customized, and strong letter from even a senior PhD grad student who mentors and knows you well is much better than a generic letter from a well-known superstar who does not know you well.

@@

Sometimes students would go to great lengths just to get letters from well-known professors in their school (e.g., department head or dean). But as mentioned, if these professors do not know you, their letters would likely be generic and carry little value (sometimes \red{red flags}). Moreover, a top professor at your university might not be well-known to US faculty (see more details in \autoref{sec:admin-letters} and \autoref{chap:your-school}). So save the trouble and get letters from \emph{any} professors/supervisors who know you well and can write a good letter about \emph{your} research ability. It's better to have a good personalized letter about your own research ability from someone who is less well-known than a generic/weak letter from a well-known person.

@@

\subsection{Self-written Letters are Bad}\label{sec:self-letters}
Many students write letters themselves, typically due to the requests of their profs. or letter writers. Such letters have \emph{little value} and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the ref. writer talking about you, in it is you who write about yourself (and they just sign the letter).

@@

\section{Asking for LORs}\label{sec:asking-lor}

As mentioned in~\autoref{sec:lor-writers}, LoR writers should be someone who knows you well and has the credibility to evaluate your research ability.  You can explicitly ask if they are willing to write a strong letter for you.  If they are not willing (you should be thankful that they are honest with you), then you should ask someone else.

% Below are some tips to approach your LoR writers:
% \begin{itemize}
% \item \textbf{Ask in advance}: You should ask for LoRs \emph{at least a month} before the deadline. People are busy with their own work and likely have many other students asking for LoRs. Also, a strong LoR takes time to write, so you should give them enough time to write a strong letter for you.

% \item \textbf{Help your writers}: You should provide your writers with details about the programs you are applying to, the deadlines, and any other information they need to know.  You should also provide them with details about your research experience and potential so that they can write a strong letter for you.
% \item \textbf{Follow up}: After you ask for LoRs, you should follow up with your writers to make sure they have submitted the letters on time.  You should also thank them for writing the letters for you.
% \item \textbf{Waive your right}: When you ask someone to write a letter for you, you will be asked if you want to waive your right to see the letter.  You should always waive your right.  This shows that you trust your writers and that you are not trying to twist their words.
% \item \textbf{Remind your writers}: After you enter your writers' information in the application system, you should tell your writers about that and let them know they will soon receive an email from the university to submit their letters.  You should also remind them when you submit your application and ask them to submit their letters on time if they haven't done so.
% \item \textbf{Help your writers}: To help your writers write a strong letter for you, you should provide them with details about your research experience and potential.  You should also provide them with a draft of your SOP so that they can see what you are saying about yourself and complement that with their own perspective.
% \item \textbf{Thank your writers}: After your writers have submitted their letters, you should thank them for writing the letters for you.  You should also let them know the outcome of your applications.
% \item \textbf{Ask for feedback}: After you have been admitted or rejected, you should ask your writers for feedback on your application.  This will help you improve your application for future applications.
% \item \textbf{Stay in touch}: After you have been admitted or rejected, you should stay in touch with your writers.  This will help you build a relationship with them and will help you in the future.
% \end{itemize}

\subsection{Waiving Your Right} When you ask someone to write a letter for you, you will be asked if you want to waive your right to see the letter.  You should \emph{always} waive your right.

@@

\section{What can you do?}\label{sec:lor-what-to-do}

\subsection{Help your LOR Writer}\label{sec:help-your-LOR-writers}

 As mentioned above in \autoref{sec:self-letters} and \autoref{sec:generic-letters}, do not write your own letter and generic letters do not give much value. Thus, to help your writer to write a strong, customized LoR, you should provide them details or unique things about yourself. For example, let them know about your GPA, research and work experience, papers (if any), or anything you want them to mention.  If the GPA in your program is highly competitive (\autoref{sec:gpa}) and they know that, remind them to talk about it in the LOR. 
You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and complement that with their own perspective.

@@

\chapter{Research Experience}\label{chap:research-experience}

@@

Here we look at publications and other research experiences that can strengthen your application.
\autoref{chap:research-opportunities} provides more information on how to find research opportunities (e.g., during your undergrad study).

\section{Publications} The most concrete evidence of research ability is having \emph{papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that you have been successfully involved in research.

@@

\begin{commentbox}[Thanh:]
    Due to academic culture, professors in Vietnam usually aim for (international) journals instead of conferences. Could you give some tips on how to know whether a journal is good (CSrankings, unfortunately, only consider conferences)?
    \tcblower
    \textbf{Vu}: One way is looking at what well-known researchers publish. For example, if you are interested in a field X, you can use CSRankings to look at active faculty in X, and then look at their websites to see what journals they publish at.
\end{commentbox}
\paragraph{What if I don't have any pubs} It is understandable that many students do not have the opportunity to publish papers. Thus, other writings, even those under submissions or rejected, would still be good and better than nothing.  Be sure to upload your papers with your application and talk about them in your statement (see \autoref{chap:sop}).

@@

\section{Work Experience} Work experience at \emph{well-known research laboratories}, such as Microsoft Research, can strengthen your
application.  Note that the emphasis here is \emph{research} places, not software development or non-research work. For example, working at a FAANG (Facebook, Amazon, Apple, Netflix, Google) company as a software engineer is not research experience, though it can be helpful for MS applications discussed in \autoref{chap:ms}. Similarly, a LoR from your supervisor for non-research experience might not count much. So do not spend much time talking about non-research job in your SOP.

Note that adcom reviewers might not be familiar with all research labs, especially those outside the US For example, while VinAI is  well-known in Vietnam and potentially in Asia, it might not be very well-known in the US. So you or your LoR writers should explicitly say something about them in your statement.  In general, if you did some good research work, then you should mention that in your SOP and ask your supervisor to write about it in their LOR (\autoref{chap:lor}).

@@

\section{Competitions} Winning \emph{internationally recognized competitions} can demonstrate your research potential.
For example, participating in Math Olympiads if you want to do theory or winning ACM programming contests if you want to ``build'' systems, e.g., software analysis.

\chapter{Statement of Purpose (SOP)}\label{chap:sop}
\chapterinfo{\emph{SOP is important}. Write it in such a way that makes you \emph{stand out}.}

@@

While you might not have control over LORs (\autoref{chap:lor}) or where your go to school (\autoref{chap:your-school}), you do over your
statement of purpose (SOP) or personal statement\footnote{Few schools separate these documents and ask you to write both: SOP, which focuses on research experiences, and personal statement, which is everything more personal, e.g., why PhD, challenges, diversity, etc}! A well-written SOP also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \autoref{sec:funding}).  Many SOP samples for CS are \href{https://cs-sop.org/}{available here}.

@@

\section{Kiss of Death in SOP} \label{sec:kiss-of-death-sop}

@@

\section{Diversity Statement}\label{sec:diversity-statement}

@@

\chapter{Your School and Grades}\label{chap:your-school}
\chapterinfo{\emph{High grades probably won't help much} (unless you're from a very top and well-known school), but bad ones likely will raise concerns.}

\section{School} Graduating from top universities \emph{that adcom members recognize helps}. For example, if you are an international student and your school is well-known, then it is \emph{``top foreign''}, which is a plus.
However, if we do not know much about schools in your country, then we are uncertain about the quality of your school and likely treat your school as \emph{``unknown foreign''}, which can be a minus point.

@@

\begin{commentbox}[Vu:]
  Sometimes PhD adcom in the US will share a document such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Foreign-Top-Schools}{this one}, which lists the top schools in several countries. We also ask other faculty and students if we think they know about the place.  For example, when I was a postdoc at UMD, members of their CS PhD adcom asked me to evaluate applicants from Vietnam.  During my time at UNL and now here at Mason, I have looked at Vietnamese applications (whether they are assigned to me or not) and provided input to their reviewers, e.g., X is the top tech school in Vietnam and so it should be \emph{top} instead of \emph{unknown} foreign, which makes a huge difference.
\end{commentbox}
\begin{commentbox}[Deepak:]
  If an applicant is anxious about their school not being known outside their country, they can provide information about their school and department, with independent sources where such information can be verified.
\end{commentbox}
\section{Grades}\label{sec:gpa}

Compared to other factors such as LoRs (\autoref{chap:lor}) and research experiences (\autoref{chap:research-experience}), grades generally do not matter much for CS PhD admission. PhD in CS is a research degree and doing well in undergrad courses does not necessarily mean you can do research (other factors such as research experiences, pubs, LoRs are more important).  In fact, many CS faculty members themselves have bad grades in undergrad courses (and some were proud of that!).

Nonetheless, if you are from a well-known school (\autoref{chap:your-school}), having good grades do help (not a lot though), e.g., adcom members often note details such as "good GPA from well-known school (\autoref{sec:ievaluate}). However if your school is not well-known, having top grades or rankings usually will not help because we cannot evaluate them (e.g., we don't know how hard it is to get a 4.0 or A's at your school). This can be an issue for students in many top international universities where the competition is so high that very good students can still have low rankings from these schools (and be overlooked by adcom).

@@

\chapter{Standard Tests (GRE and IELTS)}\label{chap:standard-tests}
\chapterinfo{\emph{Standard tests are not important}. GRE typically \emph{is not} required. For standard English tests (not required for domestic students), just do enough to pass the minimum requirements.}

\section{GREs Are Optional and Do Not Matter} While a few schools still require taking the GRE (e.g., UCF), most good CS PhD programs in the US \textbf{no longer} require it. Thus, if you haven't taken the GRE or have bad scores, then don't waste time (re)taking it. Being optional really means optional, and not taking it will not hurt your application.

@@

\section{English Tests (IELTS, TOEFL)} Unless your degrees are from specific countries such as \href{https://www.gmu.edu/international/english-language-requirements}{these}, you will need to take standardized English tests. On one hand you will need to show some level of English proficiency, but on the other hand, you do not need to have very good scores in these tests (many adcom members themselves were once international students and struggled with English).
You should just do well enough to pass the minimum requirement set by the university. %, which is required for TA (\autoref{sec:funding}).
Just as with grades (\autoref{chap:your-school}) and GRE, having high scores in these tests might not help, but having too low scores can be a \red{red flag} and sometimes results in an automatic rejection (\autoref{sec:why-rejected}), e.g., below the minimum requirement.

@@

\chapter{Miscellaneous}\label{chap:misc}
\section{CV/Resume}
\sectioninfo{Highlight and summarize major achievements, e.g., Publications, Programming Awards}

\begin{commentbox}[Vu:]
    Unlike a job application, CV is not as important for PhD admission because we do not screen applicants based on their CV.  Moreover, we care and pay attention more about your LOR, SOP, etc.  So do not spend too much time on your CV, just make sure it is easy to skim through and well-organized around research activities and achievements.
\end{commentbox}

\section{Writing Sample}\label{sec:writing-sample}
TBD TODO

\section{Online Courses and Certificates}

@@

\chapter{Interview and the Waiting Game}\label{chap:interview}
\chapterinfo{Getting an interview is typically a \emph{good sign}; but no interview does not mean rejection}

@@

\section{Interviews}

@@

\section{Not Getting Interviews}\label{sec:no-interview}
While it is generally good to get an interview, not getting one \emph{does not} mean you're out.  Many programs do not have the tradition of interviewing applicants. For example, at Mason, most admitted students with TA (\autoref{sec:ta}) do not go through interviews.

@@

\section{Notification Timeline: Why rejection letters are sent so late?}\label{sec:late-rejection}
\sectioninfo{Grad programs often wait for the accepted students to make their decisions, typically by April 15, before sending out rejection letters.}

@@

Not much you can do other than to be patient and wait. Do not send emails asking about interviews or status (unless you have interviewed specifically with someone then you can ask that person for status updates and other questions (\autoref{sec:accept-postpone-decline}).

@@

% You can then share this sheet with your reference writers and remind them to submit LoRs on your behalf (see \autoref{chap:lor}). Also, you should periodically update your writers with your status; a simple note such as \emph{``I am getting interviewed by prof. X at school Y. Do you have any advice?''} can get you a lot of help.

@@

\part{Funding}\label{part:funding}
\chapter{Funding}\label{sec:funding}
\chapterinfo{TAs, RAs, and fellowships are main funding sources for PhDs.  TAs are provided by the department to help with classes. RAs are given by profs. to help with their research.  Fellowships, provided by the university, department, or external sources such as government or industry, give move flexibility but can be very competitive.}

@@

%These questions are not useful and should be avoided. First, no one knows! Even adcom members can't tell if you would get admitted to their programs or not. As mentioned throughout this guide, the admission process is complex and involves many factors that are not part of your profile, e.g., whether the faculty have funding or are interested in your background. Second, answers to this questions are often not helpful to you.  Faculty and people knowledgeable about the process will not answer these questions, and those who do are often not knowledgeable and might give you false hope or despair. OK, someone say your GRE is low (which does not even matter for most schools (\autoref{chap:standard-tests}) , are you going to retake it?  Or someone says your GPA is low, are you going to apply to a different school?  No, you should apply to the schools you want to go to and see what happens.  If you don't get in, then you can ask for feedback and improve your profile for the next round.

@@

For each application, our system compiles a single PDF file, which consists of a summary (degrees, GPAs, etc), transcripts, test scores, LoRs, a CV, SOP, and writing samples. I usually read in this order. I start with the \emph{summary}, checking for low GPAs or test scores below the university minimum (\autoref{chap:standard-tests}). I then skim the \emph{transcripts} for low grades in relevant courses, noting issues like ``many low grades in main courses or unknown international school with good GPA''.  These are not as important as LORs or SOP but I read them first because the review system has questions about them (e.g., ``is GPA good? is IELTS sufficient?'')

@@

  \item[LoR] Letter of Recommendation. A letter written by a professor or supervisor that assesses your qualifications and potential for graduate study (\autoref{chap:lor}).

@@

  \item[SOP] Statement of Purpose is a document written by yourself to explains your research interests, background, and reasons for applying to a PhD program (\autoref{chap:sop}).

@@

  \item[TOEFL/IELTS] Test of English as a Foreign Language. A standardized test to measure the English language ability of non-native speakers.
  \item[GRE] Graduate Record Examination. A standardized test that is an admissions requirement for many graduate schools in the US. However, it is not required for most CS PhD programs (\autoref{chap:standard-tests}).

@@

\paragraph{School \autoref{chap:your-school}} Adcom already knows about your school, which is a plus (see \autoref{chap:your-school}). You are also more familiar with the US education system and culture (so hopefully you are aware of the cultural issues listed in \autoref{sec:cultural}).

\paragraph{Standard Tests \autoref{chap:standard-tests}} You do not need to take TOEFL or IELTS because you already did your undergrad (or MS) at a US university.  You might also be more comfortable communicating in English, e.g., contacting professors (\autoref{sec:contact}).

@@

\paragraph{Research Experience \autoref{chap:research-opportunities}} You might have many opportunities to do research as an undergraduate, e.g., through REUs and internships at your undergrad university.  Highlight such experience in your application.

@@

\subsection{Funding}\label{sec:ms-funding}
Unlike PhD programs, which often have funding (\autoref{sec:funding}), MS students are typically \emph{not} funded. This is because MS students do not focus on research and thus are not funded through RA and PhD students have priority for TA positions.

\begin{itemize}
\item[\textbf{RA}] Profs. are not willing to take MS students as RAs because they are not around long enough to be productive. It can take a while for a student to get used to research and start being productive, and by that time (1--2 years), they are about to graduate. Moreover, the goal of most MS students is to get a job, not to do research, so they are not as motivated to do research.

\item[\textbf{TA}] MS students are typically not given TA positions because PhDs are given priority as they will stay much longer (\autoref{sec:time}). Many depts. do not have enough TAs for PhD students and so they cannot afford to give them to MS students.
\end{itemize}

However, there are always exceptions, e.g., if you have a strong background and can demonstrate that you can be productive in research, then a prof. might be willing to take you as their RA.
Moreover, some schools have TAs for MS students, and you can apply for these positions. For example, Mason CS has quite a few TAs for MS students, and my courses in the past two years have MS TAs.

@@

\item \textbf{Living Cost:} MS students are typically not funded (\autoref{sec:ms-funding}), so living cost can be a big factor. You should consider the tuition and living expenses of the area. Note that sometimes living costs, e.g., renting, seem scary at first but students often find way to make it work, e.g., by sharing an apartment with other students.

@@

\chapter{Research Opportunities}\label{chap:research-opportunities}
\chapterinfo{How to get research experience as an undergrad.}

@@

By contributing code, fixing bugs, implementing new features, or providing documentation, students can gain practical research experience and interact with experienced developers and researchers in the field. Not only you gain research experience, but you might be able to get a LoR from the project maintainer, and you should write about this experience in your SOP (\autoref{chap:sop}.)

@@

% \chapter{Fellowship Applications}\label{sec:fellowship}
% \chapterinfo{How to apply for fellowships such as NSF GRFP, DOD DSEG to fund your PhD.}
